---
title: "An analysis of undergraduate major attrition rates within the department of computer science at CSUCI"
subtitle: "MATH 398 (Fall 2025)"
author: "Monique Calatayud, Robert McEachron, and Morgan McMurray"
format: 
  html:
    df-print: paged
    toc: true
    toc-expand: 2
    toc-location: left
    embed-resources: true
execute:
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
knitr:
  opts_chunk:
    message: false
    warning: false
---

# Introduction

Despite ongoing efforts to increase diversity in the technology sector,
computer science programs continue to be disproportionately represented
by white males. Limited minority representation within the profession
contributes to the exclusion of diverse perspectives in product design,
which can result in algorithmic harm—technology-driven threats to civil
rights and democratic values. Understanding the factors contributing to
underrepresentation in computer science is therefore critical to
fostering equitable and inclusive technological development. To the best
of our knowledge, no such analysis has been conducted on students
currently attending California State University Channel Islands (CSUCI).
Given that CSUCI is a Hispanic-Serving Institution.

Thus, the current study will examine demographics among student's
enrolled in the Computer Science Program at CSUCI, while painting a
holistic picture of the department. Additionally, factors predicting
graduation rates among students in the program will expose problem areas
necessary for intervention. The findings aim to help CSUCI and similar
institutions design targeted interventions to support at-risk computer
science students, ultimately contributing to a more diverse and
inclusive technology workforce that better serves society as a whole.

# Method

## Data

All code was analyzed using R version 4.5.2.

### Data Description

The data was obtained from the Office of Institutional Research (IR) and
included 1,091 students enrolled in the Department of Computer Science
at California State University Channel Islands (CSUCI) between Fall 2015
and Fall 2025. Fall, Winter, Spring, and Summer terms were included.
There was a total of 11,407 rows, each representing one term for a
student.

The following includes an explanation of our static variables, meaning
they do not change across the student's terms:

```         
  Student: Numerical annoymized student identifier
  Term_Code: numeric identifier for term
  Term_Desc: word description of the term that corresponds to Term_Code
  First_CI_Term: numeric identifier for the term that students started at CI
  First_CI_Term_Code: word description of First_CI_Term
  First_CS_Term: numeric identifier for the term that students started 
  First_CS_Term_Code: word description of First_CS_Term
  Did_Graduate: binary, 1 for yes, 0 for no
```

The following variables are time series dependent, meaning the values
are dynamic across each student's terms:

```         
  Time Series Dependent (values change across each student's term)
  Is_Hispanic: binary, 1 for yes, 0 for no
  Gender: Categorical, values include Man, Woman, and Non-Binary
  Major1, Major2, Major3: major(s) that students had declared during the term in question
  Major1 Descr, Major2 Descr, Major3 Descr: 
  Is_Transfer: binary, 1 for yes, 0 for no       
  Unit_Load: numeric, number of units attempted
  Lived_On_Campus: binary, 1 for yes, 0 for no
  Num_Tutoring_Visits: integer, count of visits to LRC, WMC, and STEM Center
```

### Data Cleaning

The following process was conducted to focus on non-current
undergraduate students who declared only one major throughout their
academic trajectory.

To simplify the data, term description and the major1-3 columns were
removed. The major description columns were renamed to include
underscores for R accessibility.

Multiple remaining rows per student had the same term code but different
dependent codes. The last unique term row per student was maintained;
the rest were removed (*n* = 2150 rows). Students whose last term row
did not match their graduation term code were also removed (*n* = 277
rows), as time series variables were not included in their graduation
term. Additionally, current students (*n* = 2,053 rows), students who
declared two or more majors at any term (*n* = 573 rows), and those who
were in a Master's program (*n* = 12 rows) were removed. After these
exclusions, N = 6342 term rows and N = 1091 students remained.

The Did_Graduate column was re-coded to represent students who: Did not
graduate (0), graduated with one of the majors within the computer
science program (1), or graduated with another major (0). Finally, rows
corresponding to terms outside of a student's enrollment in computer
science program were excluded (*n* = 640 rows), leaving the final
dataset for pivoting (N = 5702 rows).

```{r}
# Load necessary libraries
library(tidyverse)
library(readxl)

# Read excel data into dataframe
orig_data <- read_excel("CS Major Data 2025-10-10.xlsx", guess_max = min(8400, n_max = NULL))

# Get rid of term description columns and major1, major2, and major3 columns (we have description columns for these instead)
removed_term_descs_and_major_codes <- orig_data %>% select(-one_of("Term_Desc", "First_CI_Term_Desc", "First_CS_Term_Desc", "major1", "major2", "major3"))

# Remove duplicates, retain last of near-duplicate rows (1 row per term per student)
de_duped_reduced <- removed_term_descs_and_major_codes %>% group_by(Student, Term_Code) %>% slice_tail(n = 1) %>% ungroup()

# Rename major description columns to include underscores (makes them easier to work with)
names(de_duped_reduced)[which(colnames(de_duped_reduced) == "major1 Descr")] <- "major1_Descr"
names(de_duped_reduced)[which(colnames(de_duped_reduced) == "major2 Descr")] <- "major2_Descr"
names(de_duped_reduced)[which(colnames(de_duped_reduced) == "major3 Descr")] <- "major3_Descr"

#Remove students whose last term row doesn't match their graduation code
mismatch <- de_duped_reduced %>% 
  filter(Did_Graduate != 0) %>% 
  group_by(Student) %>% 
  slice_tail(n = 1) %>% 
  filter(Term_Code != Did_Graduate) %>% 
  pull(Student)

ddr_filtered_mismatch <- de_duped_reduced %>% 
  filter(!Student %in% mismatch)

#Identify current students and remove them
ddrfm_removed_current_students <- ddr_filtered_mismatch %>% 
  group_by(Student) %>% 
  filter(!any(Term_Code == 2258)) %>% 
  ungroup()

# Identify and keep single majors, exclude everyone else
only_single_majors <- ddrfm_removed_current_students %>% 
  group_by(Student) %>% 
  filter((all(is.na(major2_Descr)) == TRUE) && (all(is.na(major3_Descr)) == TRUE)) %>%
  ungroup()

# Remove students who graduated with a master's program in computer science
only_undergrads <- only_single_majors %>% group_by(Student) %>% filter(!any(major1_Descr == "MS: Computer Science")) %>% ungroup()

# Recode Did_Graduate to the following scheme:
# 0 --> No
# nonzero AND major1_Descr == one of: "BS: Computer Science," "BS: Information Technology," "BS: Mechatronics Engineering," "BS: Mechatronics" --> Yes, with CS/IT/EMEC
# nonzero AND major1_Decr is anything else --> Yes, with another major
cs_dept_majors <- c("BS: Computer Science", "BS: Information Technology", "BS: Mechatronics Engineering", "BS: Mechatronics")
students_wo_cs <- only_undergrads %>% 
  group_by(Student) %>% slice_tail(n = 1) %>% 
  filter(!(major1_Descr %in% cs_dept_majors)) %>% ungroup() %>% pull(Student)

recoded_graduation <- only_undergrads %>% 
  group_by(Student) %>% mutate(Did_Graduate = case_when(Did_Graduate == 0 ~ "No",
                                                        (Did_Graduate != 0) & (Student %in% students_wo_cs) ~ "Yes, with another major",
                                                        TRUE ~ "Yes, with CS/IT/EMEC")) %>% ungroup()

# Exclude all rows outside of when students were enrolled in a program under the department of computer science
only_cs_rows <- recoded_graduation %>% group_by(major1_Descr) %>% filter(major1_Descr %in% cs_dept_majors)

# Reorder columns to put static data in one area of the dataframe, and time-series in the other
cleaned_and_reordered <- only_cs_rows %>% select(Student, Is_Hispanic, Gender, 
                                                     Is_Transfer, Dependent_Income_Code, 
                                                     First_CI_Term, First_CS_Term, major1_Descr, 
                                                     major2_Descr, major3_Descr, Did_Graduate, Term_Code, 
                                                     Unit_Load, Lived_On_Campus, Num_Tutoring_Visits)
```

### Data Pivoting

The dataset was pivoted to represent one row per student in order to
ensure compatibility with statistical methods and reduce redundancy.

Because the number of terms a student was enrolled in a major from the
Computer Science Program varies by student, dataset pivoting was done
through a for loop that created new rows based on whether a change in
the "Student" column was observed. A detailed walk through of the code
is written below.

The primary statistical method in this analysis is logistical
regression. Since the original data was collected as a time series,
independence (one of the assumptions of logistic regression) was
violated. Pivoting the original dataset such that each observation is an
individual student creates independence and, in part, assures that the
assumptions for logistic regression are met.

Due to the small sample size of non-binary (*n* = 1), their value was
re-coded as being female for future analyses.

```{r}
# Get first row of all static data columns
static_data <- c("Student", "Is_Hispanic", "Gender", "Is_Transfer", 
                 "Dependent_Income_Code", "First_CI_Term", "First_CS_Term", 
                 "major1_Descr", "major2_Descr", "major3_Descr", "Did_Graduate")

pivoted <- cleaned_and_reordered[1, static_data]

# Get relevant pivoting variables from the first student
# Such that the pivoting loop can be initialized
current_id <- cleaned_and_reordered[1, "Student"]
current_major1 <- cleaned_and_reordered[1, "major1_Descr"]
current_major2 <- cleaned_and_reordered[1, "major2_Descr"]
current_major3 <- cleaned_and_reordered[1, "major3_Descr"]
current_term <- 1
current_entry <- 1

# The loop that pivots the data
for (current_row in 1:nrow(cleaned_and_reordered)) {
  
  # If the current student number is different from the saved student number,
  # Save new student number and their majors, reset term number,
  # and add new row with static data
  if (cleaned_and_reordered[current_row, "Student"] != current_id) {
    current_id <- cleaned_and_reordered[current_row, "Student"]
    current_major1 <- cleaned_and_reordered[current_row, "major1_Descr"]
    current_major2 <- cleaned_and_reordered[current_row, "major2_Descr"]
    current_major3 <- cleaned_and_reordered[current_row, "major3_Descr"]
    current_term <- 1
    current_entry <- current_entry + 1
    pivoted[current_entry, ] <- c(cleaned_and_reordered[current_row, static_data], rep(NA, each = (ncol(pivoted) - length(static_data))))
  }
  
  # Initialize the column names for the current term
  # These will be populated with values for the time-dependent variables
  current_col_label <- paste("T", current_term, "_", sep = "")
  current_cols <- c(paste(current_col_label, "Term_Code", sep=""),
                    paste(current_col_label, "Unit_Load", sep=""),
                    paste(current_col_label, "Lived_On_Campus", sep=""),
                    paste(current_col_label, "Num_Tutoring_Visits", sep=""),
                    paste(current_col_label, "major1_Descr_New", sep=""),
                    paste(current_col_label, "major2_Descr_New", sep=""),
                    paste(current_col_label, "major3_Descr_New", sep=""))
  
  # For the first 4 new columns
  # That is: "T_n_Term_Code", "T_n_Unit_Load", "T_n_Lived_On_Campus",
  # and "T_n_Num_Tutoring_Visits"
  for (ts_col in 1:4) {
    
    # Copy and paste the corresponding data from cleaned and reordered_data
    pivoted[current_entry, current_cols[ts_col]] <- cleaned_and_reordered[current_row, (length(static_data) + ts_col)]
  }
  
  # Enter a new major in the corresponding "new major" columns
  # if the student changed their major at the end of a term
  # Otherwise, fill the spot with NA
  if (current_major1 != cleaned_and_reordered[current_row, "major1_Descr"]) {
    
    # the number is the index of the "major1_Descr_New" string in current_cols
    pivoted[current_entry, current_cols[5]] <- cleaned_and_reordered[current_row, "major1_Descr"]
    current_major1 <- cleaned_and_reordered[current_row, "major1_Descr"]
  } else {
    pivoted[current_entry, current_cols[5]] <- NA
  }
  
  if (is.na(current_major2)) {
    
    # the number is the index of the "major2_Descr_New" string in current_cols
    if (!is.na(cleaned_and_reordered[current_row, "major2_Descr"])) {
       pivoted[current_entry, current_cols[6]] <- cleaned_and_reordered[current_row, "major2_Descr"]
       current_major2 <- cleaned_and_reordered[current_row, "major2_Descr"]
     } else {
       pivoted[current_entry, current_cols[6]] <- NA
     }
     
  } else {
     
     if ((is.na(cleaned_and_reordered[current_row, "major2_Descr"])) || (current_major2 != cleaned_and_reordered[current_row, "major2_Descr"])) {
       pivoted[current_entry, current_cols[6]] <- cleaned_and_reordered[current_row, "major2_Descr"]
       current_major2 <- cleaned_and_reordered[current_row, "major2_Descr"]
     } else {
       pivoted[current_entry, current_cols[6]] <- NA
     }
     
  }
   
  if (is.na(current_major3)) {
   
    # The number is the index of the "major3_Descr_New" string in current_cols
   if (!is.na(cleaned_and_reordered[current_row, "major3_Descr"])) {
      pivoted[current_entry, current_cols[7]] <- cleaned_and_reordered[current_row, "major3_Descr"]
      current_major3 <- cleaned_and_reordered[current_row, "major3_Descr"]
   } else {
      pivoted[current_entry, current_cols[7]] <- NA
   }
   
 } else {
    
    if ((is.na(cleaned_and_reordered[current_row, "major3_Descr"])) || (current_major3 != cleaned_and_reordered[current_row, "major3_Descr"])) {
      pivoted[current_entry, current_cols[7]] <- cleaned_and_reordered[current_row, "major3_Descr"]
      current_major3 <- cleaned_and_reordered[current_row, "major3_Descr"]
    } else {
      pivoted[current_entry, current_cols[7]] <- NA
    }
    
 }
  
  # Move on to the next term
  current_term <- current_term + 1
}
```

### Data Subsetting

Binning, Down-sampling Tutoring_All was created by summing up the total
number of tutoring visits the student attended across the number of
terms they were enrolled in. These numbers were categorized into No
tutoring, low (1 to 5 visits), moderate (6 to 10) and high (over 11
visits).

## Descriptive Statistics

```{r}
#| message: false
#| warning: false
#| include: false
library(tidyverse)
library(plotly)
ddata <- pivoted
```

```{r}
#| message: false
#| warning: false
#| include: false
# Removes NA Values and sums rows for each student if over 1
BinaryChecker <- function(x) { 
              x[is.na(x)] <- 0
              as.integer(rowSums(x) > 0)
}
# Removes NA Values and sums rows for each student
SummaryChecker <- function(x) { 
              x[is.na(x)] <- 0
              as.integer(rowSums(x))
}
# Recodes Gender to binary 
  ddata$Gender <- as.character(ddata$Gender)
  ddata$Gender_Binary <- recode(ddata$Gender,
                                "Man" = "0",
                                "Woman" = "1",
                                .default = "1"
  )
  
ddata$Did_Graduate <- as.character(ddata$Did_Graduate)
ddata$Did_Graduate <- recode(ddata$Did_Graduate,
                            "No" = 0,
                            "Yes, with another major" = 0,
                            "Yes, with CS/IT/EMEC" = 1,
)


#Runs Binary/Summary Checker
# ddata$Did_Graduate <- BinaryChecker(ddata[, grep("_Graduate$", names(ddata))])
ddata$Units_All <- SummaryChecker(ddata[, grep("_Unit_Load$", names(ddata))])
ddata$Housing_All <- BinaryChecker(ddata[, grep("_Lived_On_Campus$", names(ddata))])
ddata$Tutoring_All <- SummaryChecker(ddata[, grep("_Tutoring_Visits$", names(ddata))])

# Adds & Calculates Terms Attended
ddata <- ddata %>% 
        rowwise() %>% 
        mutate(Terms_Attended = sum(!is.na(c_across(ends_with("_Unit_Load"))))) %>%
        ungroup()
```

### Sample #1

Descriptive analyses were conducted to summarize the characteristics of
those in the first sample. Our sample included a total of 1,091
students, with 84.4% males (*n* = 921) and 15.6% females (*n* = 169).
Additionally, 60.9% of our sample identified as being Non-Hispanic (*n*
= 664) and 39.1% identified as being Hispanic (*n* = 427). About half of
the students did not transfer (54.1%; *n* = 590), with the other half
transferring (45.9%; *n* = 501).

In regards to students' class taking habits, the average units taken per
term was 11.42 while students stayed an average of 5.23 terms (*SD* =
3.11). Within the department, 73.9% majored in Computer Science, 16.7%
majored in Information Technology, and the remaining 4% majored in
Mechatronics Engineering. Now that we have a better understanding of our
sample, we will explain the background of our statistical analyses.

```{r}
#| column: screen-inset-shaded
#| layout-nrow: 1
# Gender Plot
ggplot(sdata, mapping = aes(x = fct_infreq(Gender_Binary), fill = Gender_Binary)) + 
      geom_bar() +
      scale_fill_manual(values= c("#E798A1","#A90806","#A90806","#E798A1"), guide="none") +
      labs(x = "Gender", y = "Number of Students") + 
      geom_text(stat = 'count', aes(label = ..count..), vjust = -0.5)

# Hispanic Plot
ggplot(sdata, mapping = aes(x = Is_Hispanic, fill = Is_Hispanic)) + 
      geom_bar() +
      labs(x = "Hispanic Staus", y = "Number of Students") +
      scale_fill_manual(values= c("#A90806","#E798A1"), guide="none") +
      scale_x_discrete(labels=c("Non-Hispanic", "Hispanic")) +
      geom_text(stat = 'count', aes(label = ..count..), vjust = -0.5)

# Transfer Plot
ggplot(sdata, mapping = aes(x = fct_infreq(Is_Transfer), fill = Is_Transfer, group = Is_Transfer)) +
      geom_bar() +
      labs(x = "Transfer Status", y = "Number of Students") +
      scale_fill_manual(values= c("#E798A1","#A90806"), guide="none") +
      scale_x_discrete(labels=c("Non-Transfer","Transfer")) +
      geom_text(stat = 'count', aes(label = ..count..), vjust = -0.5) 

# Lived on Campus
ggplot(sdata, mapping = aes(x = Housing_All, fill = Housing_All)) +
      geom_bar() +
      labs(x = "Lived On Campus (Throughout Education)", y = "Number of Students") +
      scale_fill_manual(values= c("#A90806","#E798A1"), guide="none") +
      scale_x_discrete(labels=c("Off-Campus","On-Campus")) +
      geom_text(stat = 'count', aes(label = ..count..), vjust = -0.5) 
  
# Graduation 
ggplot(sdata, mapping = aes(x = Did_Graduate_Binary, fill = Did_Graduate_Binary)) +
      geom_bar() +
      scale_fill_manual(values= c("#A90806","#E798A1"), guide="none") + 
      labs(x = "Graduation Status", y = "Number of Students") +
      geom_text(stat = 'count', aes(label = ..count..), vjust = -0.5) 

# Tutoring Binned
ggplot(sdata, mapping = aes(x = Tutoring_Binned, fill = Tutoring_Binned)) +
      geom_bar() +
      labs(x = "Student Status (Based on Units)", y = "Number of Students") +
      scale_fill_manual(values = c("#E798A1", "#A90806", "#E798A1", "#A90806"), guide = "none") +
      geom_text(stat = 'count', aes(label = ..count..), vjust = -0.5)   
  
# Student Status (Units)
ggplot(sdata, mapping = aes(x = Binned, fill = Binned)) +
      geom_bar() +
      scale_fill_manual(values= c("#E798A1", "#A90806", "#E798A1"), guide = "none") + 
      labs(x = "Student Status (Based off Units)", y = "Density",) + 
      geom_text(stat = 'count', aes(label = ..count..), vjust = -0.5)

# Units Density Plot  
ggplot(sdata, mapping = aes(x = Units_All)) +
      geom_density(adjust = 1L, fill = "#F6A6B0") +
      labs(x = "Units Taken", y = "Density",
      title = "Density of Units Taken") +
      theme(plot.title = element_text(size = 18L, hjust = 0.5),
            axis.title.y = element_text(size = 15L),
            axis.title.x = element_text(size = 15L))
```

### Sample #2

# Statistical Analyses

### Chi-Square

#### Theory

#### Assumptions

### Logistic Regression

#### Theory

#### Assumptions

A binary logistic regression was used to predict graduation.

A binary logistic regression is a multivariate prediction method that
allows for categorical and continuous variables to predict a binary
outcome. The logistic function, the sigmoid curve, models the
relationship between the predictor variables the probability of the
outcome.

Odd ratio compares the probability of the event -Include the math, which
variables used, assumptions, etc

### Survival Analysis

#### Theory

#### Assumptions

# Results

## Sample #1

### Inferential Statistics

Computer science programs have historically under-represented women and
non-binary populations. This same relationship can be seen from the data
from Channel Islands. Due to the small sample size of non-binary (*n* =
1), their value was re-coded as being female for future analyses.

```{r}
#| message: false
#| warning: false
mdata <- ddata[ddata$Gender_Binary == 0, ]
fdata <- ddata[ddata$Gender_Binary == 1, ]
```

It was found that both males and females stay at Channel Islands roughly
the same amount of terms with males staying 5.11 terms and females
staying 5.59 terms on average. A t-test determined a non-significant
difference between the two groups (*p* \> .05).

```{r}
#| eval: false
#| message: false
#| warning: false
#| paged-print: true
mean(mdata$Terms_Attended) # Male Terms Mean
mean(fdata$Terms_Attended) # Female Terms Mean
t.test(mdata$Terms_Attended, fdata$Terms_Attended, alternative = "two.sided", var.equal = FALSE, conf.level = 0.95)
```

Graduation rates were found to be non-significantly different (*p* \>
.05), with males graduating 44.76% of the time and females 42.24% of the
time across all terms.

There is a significant association between gender and the likelihood of
graduating (*p* \< .01), with fewer females graduating (n = 339)
compared to males, and more females did not graduate compared to males
(n = 1577).

```{r}
#| echo: true
t.test(mdata$Did_Graduate, fdata$Did_Graduate, alternative = "two.sided", var.equal = FALSE, conf.level = 0.95)
chi_gender_grad <- table(ddata$Gender_Binary, ddata$Did_Graduate)
chi_gender_grad_result <- chisq.test(chi_gender_grad)
```

#### Hispanic Status

Due to Channel Islands being a Hispanic Serving Institution (HSI),
exploration of student's Hispanic Status was important to us. The
following chart showcases the makeup of Hispanic and Non-Hispanic
student's in the Computer Science Department.

```{r}
#| message: false
#| warning: false
nhisdata <- ddata[ddata$Is_Hispanic == 0, ]
hisdata <- ddata[ddata$Is_Hispanic == 1, ]
```

Student's from both identity statuses attended same amount of terms
(*Winter, Fall, Spring, Summer*) with non-Hispanic students attending a
mean of 5.638 terms and Hispanic students 5.370. A two sided t-test
determined no significant difference between the two groups (*p* \>
.05).

```{r}
#| eval: false
mean(nhisdata$Terms_Attended) # Non-Hispanic Terms Mean
mean(hisdata$Terms_Attended) # Hispanic Terms Mean
t.test(hisdata$Terms_Attended, nhisdata$Terms_Attended, alternative = "two.sided", var.equal = FALSE, conf.level = 0.95)
```

Hispanic student's were found to graduate 37.5% of the time, whereas
their Non-Hispanic counterparts graduated 49.3% of the time. Further
examination revealed a significant difference in graduation rate among
the two groups, *t*(1449)= 4.73, *p* \< .001.

```{r}
#| eval: false    
summary(nhisdata$Did_Graduate) # Non-Hispanic Graduation Mean
summary(hisdata$Did_Graduate) # Hispanic Graduation Mean
t.test(nhisdata$Did_Graduate, hisdata$Did_Graduate, alternative = "two.sided", var.equal = FALSE, conf.level = 0.95)
```

The intersection of the aforementioned findings indicate that Hispanic
and Non-Hispanic students attend the same amount of terms, but Hispanic
students graduate at a significantly lower rate.

#### Transfer Status

Channel Islands receives lots of transfer student's due to programs in
place at the local community colleges and the institutions relatively
low cost. This is supported by the data we received below.

```{r}
#| message: false
#| warning: false
hdata <- ddata[ddata$Is_Transfer == 0, ]
tdata <- ddata[ddata$Is_Transfer == 1, ]
```

It was found that student's who started their academic journey at
Channel Islands attend more terms (*M* = 5.96) as those who transfer
(*M* = 5.15) with a significant difference, *t*(1282) = 4.957, *p* \<
.001. Interestingly, this difference was not as large as expected as
those who transfer should theoretically need far less classes then those
who start at the institution.

```{r}
#| message: false
#| warning: false
#| eval: false
summary(hdata$Terms_Attended) # Home Terms Mean
summary(tdata$Terms_Attended) # Transfer Terms Mean
t.test(tdata$Terms_Attended, hdata$Terms_Attended, alternative = "two.sided", var.equal = FALSE, conf.level = 0.95)
```

When taking into account graduation rates, the lack of disparity between
terms attended for transfer and non-transfer student becomes apparent.
Student's who started at Channel Islands graduate 35.61% of the time
whereas those who transfer graduate 51.98% of the time. A t-test
determined that this was a significant difference, *t*(1552) = 6.61, *p*
\> .001.

```{r}
#| eval: false
#| message: false
#| warning: false
summary(hdata$Did_Graduate) # Home Graduation Mean
summary(tdata$Did_Graduate) # Transfer Graduation Mean
t.test(tdata$Did_Graduate, hdata$Did_Graduate, alternative = "two.sided", var.equal = FALSE, conf.level = 0.95)
```

#### Majors

The Computer Science Department at Channel Islands includes the majors
of Computer Science, Mechatronics and Information Technology. Due to
some student's double majoring or changing programs the data includes
majors that are outside this category.

Our major count was:

```{r}
#| echo: false
#| message: false
#| warning: false
dfmajor <- as.data.frame(c(ddata$major1_Descr, ddata$major2_Descr, ddata$major3_Descr)) # Combines student's majors into a DF
dfmajor <- dfmajor %>% # Removes NA Values
            na.omit(mutate(dfmajor = str_trim(c(ddata$major1, ddata$major2, ddata$major3))))
colnames(dfmajor) <- c("Major") # Renames Column
dfmajor <- dfmajor %>% # Creates count of majors
            count(Major)
print(dfmajor)
dfmajor <- dfmajor[dfmajor$n > 20, ] # Removes major with under 20 students
```

Due to the small sample size, majors with under 20 student's were
excluded for the following chart:

#### Gender and Hispanic Status

It was found that Hispanic males graduate **39.44%** of the time,
whereas their Non-Hispanic male counterparts graduate **48.54%** of the
time. Similarly, Hispanic females graduate **29.13%** of the time and
Non-Hispanic females graduate **53.33%** of the time. This showcases a
large disparity between Hispanic and Non-Hispanic students. In addition,
there is a **\~10%** difference in graduation rates among Hispanic males
and females, whereas this disparity is not seen in their Non-Hispanic
counterparts.s

```{r}
#| eval: false
#| message: false
#| warning: false
  mhisdata <- mdata[mdata$Is_Hispanic == 1, ] # Male Hispanic 
  fhisdata <- fdata[fdata$Is_Hispanic == 1, ] # Female Hispanic
  
  mnhisdata <- mdata[mdata$Is_Hispanic == 0, ] # Male Not Hispanic
  fnhisdata <- fdata[fdata$Is_Hispanic == 0, ] # Female Not Hispanic
  
  # Gender by Hispanic Status
    summary(mhisdata$Did_Graduate) # Male Hispanic 0.3944
    summary(fhisdata$Did_Graduate) # Female Hispanic 0.2913
    
    summary(mnhisdata$Did_Graduate) # Male Not Hispanic: 0.4854
    summary(fnhisdata$Did_Graduate) # Female Not Hispanic: 0.5333
```

### Logistic Regression

-   Add a table

-   First, we need to examine what assumptions blah blah (charts and
    graphs to show linearity etc.)

-   Will include information from the logistic regression aka results

-   Will include interpretation of the results

-   Will include any charts regarding what it looks like (i.e. line of
    best fit and assumptions)

### Survival Analysis

## Sample #2

### Inferential Statistics

#### Hispanic Status

#### Transfer Status

#### Majors

### Logistic Regression

# Discussion

Results (colorful language), Possible intervention demographics,
limitations, and future directions.

# Other

## References

## Acknowledgements

"Sometimes, you end up gettin' where you wanna go, And then find out
when you get there that you don't wanna be where you're at. And then you
feel like this is all just a waste of time. But just remember, if you
never got there then you'd always be wanting to get there And then you
don't know you don't wanna be there until you get there So sometimes you
just gotta go where you don't wanna go Just to know that that's not
whеre you wanna be Okay, have a supеr freakin' awesome d— rest of your
day and stuff Okay, bye" - Rodney Norman

```{r}
#                      /^--^\     /^--^\     /^--^\
#                      \____/     \____/     \____/
#                     /      \   /      \   /      \
#                    |        | |        | |        |
#                     \__  __/   \__  __/   \__  __/
#|^|^|^|^|^|^|^|^|^|^|^|^\ \^|^|^|^/ /^|^|^|^|^\ \^|^|^|^|^|^|^|^|^|^|^|^|
#| | | | | | | | | | | | |\ \| | |/ /| | | | | | \ \ | | | | | | | | | | |
#| | | | | | | | | | | | / / | | |\ \| | | | | |/ /| | | | | | | | | | | |
#| | | | | | | | | | | | \/| | | | \/| | | | | |\/ | | | | | | | | | | | |
##########################################################################
#| | | | | | | | | | | | | | | | | | | | | | | | | | | | | | | | | | | | |
#| | | | | | | | | | | | | | | | | | | | | | | | | | | | | | | | | | | | |
```
