---
title: "An analysis of undergraduate major attrition rates within the department of computer science at CSUCI"
subtitle: "MATH 398 (Fall 2025)"
author: "Monique Calatayud, Robert McEachron, and Morgan McMurray"
format: 
  html:
    df-print: paged
    toc: true
    toc-expand: 2
    toc-location: left
execute:
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

# Introduction

Despite ongoing efforts to increase diversity in the technology sector,
computer science programs continue to be disproportionately represented
by white males. Limited minority representation within the profession
contributes to the exclusion of diverse perspectives in product design,
which can result in algorithmic harm—technology-driven threats to civil
rights and democratic values. Understanding the factors contributing to
underrepresentation in computer science is therefore critical to
fostering equitable and inclusive technological development. To the best
of our knowledge, no such analysis has been conducted on students
currently attending California State University Channel Islands (CSUCI).
Given that CSUCI is a Hispanic-Serving Institution.

Thus, the current study will examine demographics among student's
enrolled in the Computer Science Program at CSUCI, while painting a
holistic picture of the department. Additionally, factors predicting
graduation rates among students in the program will expose problem areas
necessary for intervention. The findings aim to help CSUCI and similar
institutions design targeted interventions to support at-risk computer
science students, ultimately contributing to a more diverse and
inclusive technology workforce that better serves society as a whole.

# Method

## Data

All code was analyzed using R version 4.5.2.

### Data Description

```{r}
#| message: false
#| warning: false
#| include: false
library(tidyverse)
library(plotly)
ddata <- read.csv("DeDuped_Data.csv")
```

```{r}
#| message: false
#| warning: false
#| include: false
# Removes NA Values and sums rows for each student if over 1
BinaryChecker <- function(x) { 
              x[is.na(x)] <- 0
              as.integer(rowSums(x) > 0)
}
# Removes NA Values and sums rows for each student
SummaryChecker <- function(x) { 
              x[is.na(x)] <- 0
              as.integer(rowSums(x))
}
# Recodes Gender to binary 
  ddata$Gender <- as.character(ddata$Gender)
  ddata$Gender_Binary <- recode(ddata$Gender,
                                "Man" = "0",
                                "Woman" = "1",
                                .default = "1"
  )
  
ddata$Did_Graduate <- as.character(ddata$Did_Graduate)
ddata$Did_Graduate <- recode(ddata$Did_Graduate,
                            "No" = 0,
                            "Yes, with another major" = 0,
                            "Yes, with CS/IT/EMEC" = 1,
)


#Runs Binary/Summary Checker
# ddata$Did_Graduate <- BinaryChecker(ddata[, grep("_Graduate$", names(ddata))])
ddata$Units_All <- SummaryChecker(ddata[, grep("_Unit_Load$", names(ddata))])
ddata$Housing_All <- BinaryChecker(ddata[, grep("_Lived_On_Campus$", names(ddata))])
ddata$Tutoring_All <- SummaryChecker(ddata[, grep("_Tutoring_Visits$", names(ddata))])

# Adds & Calculates Terms Attended
ddata <- ddata %>% 
        rowwise() %>% 
        mutate(Terms_Attended = sum(!is.na(c_across(ends_with("_Unit_Load"))))) %>%
        ungroup()
```

The data was obtained from the Office of Institutional Research (IR) and
included 1,091 students enrolled in the Department of Computer Science
at California State University Channel Islands (CSUCI) between Fall 2015
and Fall 2025. Fall, Winter, Spring, and Summer terms were included.
There was a total of 11,407 rows, each representing one term for a
student.

The following includes an explanation of our static variables, meaning
they do not change across the student's terms:

```         
  Student: Numerical annoymized student identifier
  Term_Code: numeric identifier for term
  Term_Desc: word description of the term that corresponds to Term_Code
  First_CI_Term: numeric identifier for the term that students started at CI
  First_CI_Term_Code: word description of First_CI_Term
  First_CS_Term: numeric identifier for the term that students started 
  First_CS_Term_Code: word description of First_CS_Term
  Did_Graduate: binary, 1 for yes, 0 for no
```

The following variables are time series dependent, meaning the values
are dynamic across each student's terms:

```         
  Time Series Dependent (values change across each student's term)
  Is_Hispanic: binary, 1 for yes, 0 for no
  Gender: Categorical, values include Man, Woman, and Non-Binary
  Major1, Major2, Major3: major(s) that students had declared during the term in question
  Major1 Descr, Major2 Descr, Major3 Descr: 
  Is_Transfer: binary, 1 for yes, 0 for no       
  Unit_Load: numeric, number of units attempted
  Lived_On_Campus: binary, 1 for yes, 0 for no
  Num_Tutoring_Visits: integer, count of visits to LRC, WMC, and STEM Center
```

### Data Cleaning

The following process was conducted to focus on non-current undergraduate students who declared only one major throughout their academic trajectory.

To simplify the data, term description and the major1-3 columns were removed. The major description columns were renamed to include underscores for R accessibility.

Multiple remaining rows per student had the same term code but different dependent codes.  The last unique term row per student was maintained; the rest were removed (*n* = 2150 rows). Students whose last term row did not match their graduation term code were also removed (*n* = 277 rows), as time series variables were not included in their graduation term. Additionally, current students (*n* = 2,053 rows), students who declared two or more majors at any term (*n* = 573 rows), and those who were in a Master's program (*n* = 12 rows) were removed. After these exclusions, N = 6342 term rows and N = 1091 students remained.

The Did_Graduate column was re-coded to represent students who: Did not graduate (0), graduated with one of the majors within the computer science program (1), or graduated with another major (0). Finally, rows corresponding to terms outside of a student's enrollment in computer science program were excluded (*n* = 640 rows), leaving the final dataset for pivoting (N = 5702 rows).

### Data Transformation

The dataset was pivoted to represent one row per student in order to ensure compatibility with statistical methods and reduce redundancy.

The pivoting algorithm works as follows:

    1. Static variable names were gathered into a vector.
    2. The final pivoted dataset was initialized with all static variables from the first row of the fully cleaned dataset.
    3. Pivoting variables []

-   Will include information about why this allows us to conduct
    analyses

Include the binary input and adding more columns Due to the small sample size of non-binary (*n* = 1), their value was re-coded as being female for future analyses.

## Descriptive Statistics

Descriptive analyses were conducted to summarize the characteristics of
those in the sample. Our sample included a total of 1,091 students, with
84.4% males (*n* = 921) and 15.6% females (*n* = 169). Additionally,
60.9% of our sample identified as being Non-Hispanic (*n* = 664) and
39.1% identified as being Hispanic (*n* = 427). About half of the
students did not transfer (54.1%; *n* = 590), with the other half
transferring (45.9%; *n* = 501).

In regards to students' class taking habits, the average units taken per
term was 11.42 while students stayed an average of 5.23 terms (*SD* =
3.11). Within the department, 73.9% majored in Computer Science, 16.7%
majored in Information Technology, and the remaining 4% majored in
Mechatronics Engineering. Now that we have a better understanding of our
sample, we will explain the background of our statistical analyses.

# Statistical Analyses

### Inferential Statistics

### Logistic Regression

A binary logistic regression was used to predict graduation.

A binary logistic regression is a multivariate prediction method that
allows for categorical and continuous variables to predict a binary
outcome. The logistic function, the sigmoid curve, models the
relationship between the predictor variables the probability of the
outcome.

Odd ratio compares the probability of the event -Include the math, which
variables used, assumptions, etc

# Results

## Inferential Statistics

### Gender

Computer science programs have historically under-represented women and
non-binary populations. This same relationship can be seen from the data
from Channel Islands. Due to the small sample size of non-binary (*n* =
1), their value was re-coded as being female for future analyses.

```{r}
#| echo: false
#| warning: false
  ggplot(data = ddata,
        mapping = aes(x = fct_infreq(Gender), fill = Gender)) + 
        geom_bar() +
        scale_fill_manual(values= c("#E798A1","#A90806","#A90806","#E798A1"), guide="none") +
        labs(x = "Gender",
             y = "Number of Students") + 
        geom_text(stat = 'count', aes(label = ..count..), vjust = -0.5)
```

```{r}
#| message: false
#| warning: false
mdata <- ddata[ddata$Gender_Binary == 0, ]
fdata <- ddata[ddata$Gender_Binary == 1, ]
```

It was found that both males and females stay at Channel Islands roughly
the same amount of terms with males staying 5.11 terms and females
staying 5.59 terms on average. A t-test determined a non-significant
difference between the two groups (*p* \> .05).

```{r}
#| message: false
#| eval: false
#| warning: false
mean(mdata$Terms_Attended) # Male Terms Mean
mean(fdata$Terms_Attended) # Female Terms Mean
t.test(mdata$Terms_Attended, fdata$Terms_Attended, alternative = "two.sided", var.equal = FALSE, conf.level = 0.95)
```

Graduation rates were found to be non-significantly different (*p* \>
.05), with males graduating 44.76% of the time and females 42.24% of the
time across all terms.

```{r}
#| echo: true
mean(mdata$Did_Graduate) # Male Graduation Mean
mean(fdata$Did_Graduate) # Female Graduation Mean
t.test(mdata$Did_Graduate, fdata$Did_Graduate, alternative = "two.sided", var.equal = FALSE, conf.level = 0.95)
```

### Hispanic Status

Due to Channel Islands being a Hispanic Serving Institution (HSI),
exploration of student's Hispanic Status was important to us. The
following chart showcases the makeup of Hispanic and Non-Hispanic
student's in the Computer Science Department.

```{r}
#| echo: false
#| warning: false
# Hispanic Plot
  dfhis <- as.factor(ddata$Is_Hispanic) # Converts to readable format for ggplot
  dfhis <- as.data.frame(dfhis)
  ggplot(data = dfhis,
        mapping = aes(x = dfhis, fill = dfhis)) + 
        geom_bar() +
        labs(x = "Hispanic Staus",
             y = "Number of Students") +
        scale_fill_manual(values= c("#A90806","#E798A1"), guide="none") +
        geom_text(stat = 'count', aes(label = after_stat(count)), vjust = -0.5) +
        scale_x_discrete(labels=c("Non-Hispanic", "Hispanic"))
```

```{r}
#| message: false
#| warning: false
nhisdata <- ddata[ddata$Is_Hispanic == 0, ]
hisdata <- ddata[ddata$Is_Hispanic == 1, ]
```

Student's from both identity statuses attended same amount of terms
(*Winter, Fall, Spring, Summer*) with non-Hispanic students attending a
mean of 5.638 terms and Hispanic students 5.370. A two sided t-test
determined no significant difference between the two groups (*p* \>
.05).

```{r}
#| eval: false
mean(nhisdata$Terms_Attended) # Non-Hispanic Terms Mean
mean(hisdata$Terms_Attended) # Hispanic Terms Mean
t.test(hisdata$Terms_Attended, nhisdata$Terms_Attended, alternative = "two.sided", var.equal = FALSE, conf.level = 0.95)
```

Hispanic student's were found to graduate 37.5% of the time, whereas
their Non-Hispanic counterparts graduated 49.3% of the time. Further
examination revealed a significant difference in graduation rate among
the two groups, *t*(1449)= 4.73, *p* \< .001.

```{r}
#| eval: false    
summary(nhisdata$Did_Graduate) # Non-Hispanic Graduation Mean
summary(hisdata$Did_Graduate) # Hispanic Graduation Mean
t.test(nhisdata$Did_Graduate, hisdata$Did_Graduate, alternative = "two.sided", var.equal = FALSE, conf.level = 0.95)
```

The intersection of the aforementioned findings indicate that Hispanic
and Non-Hispanic students attend the same amount of terms, but Hispanic
students graduate at a significantly lower rate.

### Transfer Status

Channel Islands receives lots of transfer student's due to programs in
place at the local community colleges and the institutions relatively
low cost. This is supported by the data we received below.

```{r}
#| echo: false
#| warning: false
# Transfer Plot
  dftrans <- as.factor(ddata$Is_Transfer) # Converts to readable format for ggplot
  dftrans <- as.data.frame(dftrans)
  ggplot(data = dftrans,
        mapping = aes(x = fct_infreq(dftrans), fill = dftrans, group = dftrans)) +
        geom_bar() +
        labs(x = "Transfer Status",
             y = "Number of Students") +
        scale_fill_manual(values= c("#E798A1","#A90806"), guide="none") +
        scale_x_discrete(labels=c("Non-Transfer","Transfer")) +
        geom_text(stat = 'count', aes(label = ..count..), vjust = -0.5)
```

```{r}
#| message: false
#| warning: false
hdata <- ddata[ddata$Is_Transfer == 0, ]
tdata <- ddata[ddata$Is_Transfer == 1, ]
```

It was found that student's who started their academic journey at
Channel Islands attend more terms (*M* = 5.96) as those who transfer
(*M* = 5.15) with a significant difference, *t*(1282) = 4.957, *p* \<
.001. Interestingly, this difference was not as large as expected as
those who transfer should theoretically need far less classes then those
who start at the institution.

```{r}
#| message: false
#| warning: false
#| eval: false
summary(hdata$Terms_Attended) # Home Terms Mean
summary(tdata$Terms_Attended) # Transfer Terms Mean
t.test(tdata$Terms_Attended, hdata$Terms_Attended, alternative = "two.sided", var.equal = FALSE, conf.level = 0.95)
```

When taking into account graduation rates, the lack of disparity between
terms attended for transfer and non-transfer student becomes apparent.
Student's who started at Channel Islands graduate 35.61% of the time
whereas those who transfer graduate 51.98% of the time. A t-test
determined that this was a significant difference, *t*(1552) = 6.61, *p*
\> .001.

```{r}
#| eval: false
#| message: false
#| warning: false
summary(hdata$Did_Graduate) # Home Graduation Mean
summary(tdata$Did_Graduate) # Transfer Graduation Mean
t.test(tdata$Did_Graduate, hdata$Did_Graduate, alternative = "two.sided", var.equal = FALSE, conf.level = 0.95)
```

### Majors

The Computer Science Department at Channel Islands includes the majors
of Computer Science, Mechatronics and Information Technology. Due to
some student's double majoring or changing programs the data includes
majors that are outside this category.

Our major count was:

```{r}
#| echo: false
#| message: false
#| warning: false
dfmajor <- as.data.frame(c(ddata$major1_Descr, ddata$major2_Descr, ddata$major3_Descr)) # Combines student's majors into a DF
dfmajor <- dfmajor %>% # Removes NA Values
            na.omit(mutate(dfmajor = str_trim(c(ddata$major1, ddata$major2, ddata$major3))))
colnames(dfmajor) <- c("Major") # Renames Column
dfmajor <- dfmajor %>% # Creates count of majors
            count(Major)
print(dfmajor)
dfmajor <- dfmajor[dfmajor$n > 20, ] # Removes major with under 20 students
```

Due to the small sample size, majors with under 20 student's were
excluded for the following chart:

```{r}
#| echo: false
#| message: false
#| warning: false
ggplot(data = dfmajor,
        mapping = aes(x = Major, y = n, fill = Major, group = Major)) + 
        geom_bar(stat="identity") + 
        scale_fill_manual(values= c("#A90806","#E798A1", "#A90806","#E798A1"), guide="none") +
        geom_text(stat = "Identity", aes(label = n), vjust = -0.5) + 
        labs(y = "Number of Students")
```

### Gender and Hispanic Status

It was found that Hispanic males graduate **39.44%** of the time,
whereas their Non-Hispanic male counterparts graduate **48.54%** of the
time. Similarly, Hispanic females graduate **29.13%** of the time and
Non-Hispanic females graduate **53.33%** of the time. This showcases a
large disparity between Hispanic and Non-Hispanic students. In addition,
there is a **\~10%** difference in graduation rates among Hispanic males
and females, whereas this disparity is not seen in their Non-Hispanic
counterparts.

```{r}
#| eval: false
#| message: false
#| warning: false
  mhisdata <- mdata[mdata$Is_Hispanic == 1, ] # Male Hispanic 
  fhisdata <- fdata[fdata$Is_Hispanic == 1, ] # Female Hispanic
  
  mnhisdata <- mdata[mdata$Is_Hispanic == 0, ] # Male Not Hispanic
  fnhisdata <- fdata[fdata$Is_Hispanic == 0, ] # Female Not Hispanic
  
  # Gender by Hispanic Status
    summary(mhisdata$Did_Graduate) # Male Hispanic 0.3944
    summary(fhisdata$Did_Graduate) # Female Hispanic 0.2913
    
    summary(mnhisdata$Did_Graduate) # Male Not Hispanic: 0.4854
    summary(fnhisdata$Did_Graduate) # Female Not Hispanic: 0.5333
```

## Logistic Regression

-   Add a table

-   First, we need to examine what assumptions blah blah (charts and
    graphs to show linearity etc.)

-   Will include information from the logistic regression aka results

-   Will include interpretation of the results

-   Will include any charts regarding what it looks like (i.e. line of
    best fit and assumptions)

# Discussion
